---
title: "Reduce Dimensions"
author: "Ben Southgate"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: textmate
    df_print: paged
editor_options:
  chunk_output_type: console
---
  
# Setup
  
Set chunk options:
  
```{r}
knitr::opts_chunk$set(echo = TRUE, dev = "png", fig.align = "center")
```

Set library path:
  
```{r}
.libPaths(new = "/sbgenomics/project-files/library")
```

Load Bioconductor packages:
  
```{r, warning=FALSE, message=FALSE}
library(scater)
library(scran)
library(scuttle)
library(Seurat)
```

Load CRAN packages:
  
```{r, warning=FALSE, message=FALSE}
library(patchwork)
library(ggplot2)
library(dplyr)
library(reshape2)
library(viridis)
library(plyr)
```

Source user-defined functions:
  
```{r, warning=FALSE, message=FALSE}
source("scripts/reduceDims.R")
```

Enable multicore parallel evaluation:
  
```{r, warning=FALSE, message=FALSE}
library(BiocParallel)

BPP <- MulticoreParam(
  workers = 6,
  stop.on.error = TRUE,
  RNGseed = 0305

)
```

# Processing

Read experiment object:
  
```{r, warning=FALSE, message=FALSE}
seurat <- readRDS("/sbgenomics/output-files/03-normalisation.rds")
```

## PCA

PCA individual samples

```{r, warning=FALSE, message=FALSE}
seurat <- bplapply(seurat, pca, BPPARAM = BPP)
```

### Plot PCAs

Individual PCA - dimplot

```{r, warning=FALSE, message=FALSE}
pca_plots_dim <- lapply(seurat, dimplot, reduction = "pca", label = FALSE, group.by = "Group")
```

Wrap plot PCAs dimplots

```{r, fig.width = 10, fig.height = 10}
wrap_plots(pca_plots_dim, ncol = 2)
```

Individual PCA - featureplot ncount

```{r, warning=FALSE, message=FALSE}
pca_plots_feature <- lapply(seurat, featureplot, reduction = "pca", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot PCAs featureplot ncount

```{r, fig.width = 10, fig.height = 10}
wrap_plots(pca_plots_feature, ncol = 2)
```

Individual PCA - featureplot nfeature

```{r, warning=FALSE, message=FALSE}
pca_plots_feature <- lapply(seurat, featureplot, reduction = "pca", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot PCAs featureplot nfeature

```{r, fig.width = 10, fig.height = 10}
wrap_plots(pca_plots_feature, ncol = 2)
```

## Finding dimensionality of dataset

Find ncomponents for downstream dimensionality reduction

### Elbow plot

```{r, warning=FALSE, message=FALSE}
elbows <- lapply(seurat, elbow)
```

```{r, fig.width = 8, fig.height = 10}
wrap_plots(elbows, ncol = 2)
```

Take number of PCs from visual inspection - better to go higher

Define elbow vector:

```{r, warning=FALSE}
elbow_vector <- c("1:10", "1:10", "1:10", "1:10", "1:10", "1:10")

names(elbow_vector) <- names(seurat)

print(elbow_vector)
```

Add PC elbow to object

```{r, warning=FALSE, message=FALSE}
seurat <- lapply(seurat, addElbow, elbow_vector = elbow_vector)
```

## UMAP

### Test stability of UMAP params
n.neighbors should be in the range 5 to 50.
min.dist sensible values are in the range 0.001 to 0.5.
```{r, Explore-Features-UMAP, warning=FALSE, message=FALSE}
seurat_1 <- bplapply(seurat, umap, n.neighbors = 30, min.dist = 0.3, BPPARAM = BPP)

seurat_2 <- bplapply(seurat, umap, n.neighbors = 5, min.dist = 0.25, BPPARAM = BPP)

seurat_3 <- bplapply(seurat, umap, n.neighbors = 10, min.dist = 0.1, BPPARAM = BPP)

seurat_4 <- bplapply(seurat, umap, n.neighbors = 20, min.dist = 0.1, BPPARAM = BPP)

seurat_5 <- bplapply(seurat, umap, n.neighbors = 10, min.dist = 0.001, BPPARAM = BPP)
```

UMAP n.neighbors = 30, min.dist = 0.3
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_1, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 5, min.dist = 0.25
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_2, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 10, min.dist = 0.1
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_3, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 20, min.dist = 0.1
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_4, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 10, min.dist = 0.001
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_5, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

Clean up workspace
```{r}
rm(seurat_1, seurat_2, seurat_3, seurat_4, seurat_5)
gc()
```

Run UMAP individual samples for choice of params (standard seems fine)

```{r, warning=FALSE, message=FALSE}
seurat <- bplapply(seurat, umap, BPPARAM = BPP)
```

### Plot UMAP

Individual UMAPs

```{r, warning=FALSE, message=FALSE}
umap_plots <- lapply(seurat, dimplot, reduction = "umap", label = FALSE, group.by = "Group")
```

Wrap plot UMAP

```{r, fig.width = 10, fig.height = 10}
wrap_plots(umap_plots, ncol = 2)
```

Individual UMAP - featureplot ncounts

```{r, warning=FALSE, message=FALSE}
umap_plots_feature <- lapply(seurat, featureplot, reduction = "umap", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot UMAP featureplot ncounts

```{r, fig.width = 10, fig.height = 10}
wrap_plots(umap_plots_feature, ncol = 2)
```

Individual UMAP - featureplot nfeatures

```{r, warning=FALSE, message=FALSE}
umap_plots_feature <- lapply(seurat, featureplot, reduction = "umap", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot UMAP featureplot nfeatures

```{r, fig.width = 10, fig.height = 10}
wrap_plots(umap_plots_feature, ncol = 2)
```

## TSNE

### Test stability of TSNE params

perplexity parameter (should not be bigger than 3 * perplexity < nrow(X) - 1)
NB: Try N^1/2 as rule of thumb

```{r, Explore-Features-TSNE-N1-2, warning=FALSE, message=FALSE}
seurat_1 <- bplapply(seurat, tsne, perplexity = NULL, BPPARAM = BPP)
```

```{r, Explore-Features-TSNE, warning=FALSE, message=FALSE}
seurat_2 <- bplapply(seurat, tsne, perplexity = 5, BPPARAM = BPP)

seurat_3 <- bplapply(seurat, tsne, perplexity = 10, BPPARAM = BPP)

seurat_4 <- bplapply(seurat, tsne, perplexity = 20, BPPARAM = BPP)

seurat_5 <- bplapply(seurat, tsne, perplexity = 100, BPPARAM = BPP)
```

perplexity = N^1/2
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_1, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 5
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_2, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 10
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_3, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 20
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_4, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 100
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_5, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

Clean up workspace
```{r}
rm(seurat_1, seurat_2, seurat_3, seurat_4, seurat_5)
gc()
```

Run tsne individual samples for choice of params (perplexity = N^1/2)

```{r, warning=FALSE, message=FALSE}
seurat <- bplapply(seurat, tsne, perplexity = NULL, BPPARAM = BPP)
```

### Plot TSNE

Individual TSNEs

```{r, warning=FALSE, message=FALSE}
tsne_plots <- lapply(seurat, dimplot, reduction = "tsne", label = FALSE, group.by = "Group")
```

Wrap plot TSNE

```{r, fig.width = 10, fig.height = 10}
wrap_plots(tsne_plots, ncol = 2)
```

Individual TSNE - featureplot ncounts

```{r, warning=FALSE, message=FALSE}
tsne_plots_feature <- lapply(seurat, featureplot, reduction = "tsne", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot TSNE featureplot ncounts

```{r, fig.width = 10, fig.height = 10}
wrap_plots(tsne_plots_feature, ncol = 2)
```

Individual TSNE - featureplot nfeatures

```{r, warning=FALSE, message=FALSE}
tsne_plots_feature <- lapply(seurat, featureplot, reduction = "tsne", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot TSNE featureplot nfeatures

```{r, fig.width = 10, fig.height = 10}
wrap_plots(tsne_plots_feature, ncol = 2)
```

# Output

Save experiment object:

```{r}
saveRDS(seurat, file = "/sbgenomics/output-files/04-reduced-dimensions.rds")
```  

Print session information:
  
```{r}
sessionInfo()
```


