---
title: "Normalisation"
params:
  nfeatures:
    - 2000
  min_cells:
    - 2
  example:
    - "Foxj1"
---
  
## Setup

Add child documents

```{r}
#| child = "analysis/big-data.qmd"
```

Restore project dependencies:

```{r}
#renv::restore()
```

Load Bioconductor packages:
  
```{r, message=FALSE}
library(Seurat)
```

Load CRAN packages:
  
```{r, message=FALSE}
library(patchwork)
library(ggplot2)
library(dplyr)
library(reshape2)
```

Source user-defined functions:

```{r}
source("scripts/normalise.R")
```

## Processing

Read experiment object:
  
```{r}
seurat <- readRDS("output/02-quality-control.rds")
```

### Feature selection

#### VST approach

Top n HVGs

```{r, message = FALSE, warning = FALSE, Find-features-n}
features_n <- bplapply(seurat, findFeatures, nfeatures = params$nfeatures, BPPARAM = BPPARAM) 
```

Feature plot n features

```{r, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5, warning=FALSE}
wrap_plots(lapply(features_n, plotVariableGenes, selection.method = "vst"), ncol = 2)
```

### Normalisation

Using SCT transform

See https://satijalab.org/seurat/reference/sctransform for more details.

Seurat authors advice each sample to be normalised separately prior to merging
See https://github.com/satijalab/seurat/discussions/4430 GitHub for more details.

Here we normalise each sample alone and do not merge.

Set min cells as per params

```{r, Normalise-SCT-features-n, message = FALSE, warning=FALSE}
seurat <- lapply(seurat, normalise, variable.features.n = params$nfeatures, min_cells = params$min_cells) 
```

Plot detailed steps of sctransform

Details of sct transform from sctransform::vst vignette
See https://htmlpreview.github.io/?https://github.com/satijalab/sctransform/blob/supp_html/supplement/variance_stabilizing_transformation.html for more details

```{r, message = FALSE, warning=FALSE}
vst_out <- lapply(seurat, normaliseSctransform) 
```

Plot model params

```{r, fig.width = 10, fig.height = 5, message = FALSE, warning=FALSE}
wrap_plots(lapply(vst_out, sctransform::plot_model_pars, show_theta = TRUE))
```

Plot model for example gene 

```{r, message = TRUE, warning=FALSE, fig.show="hide"}
print(params$example)
modl <- lapply(seurat, plotModelSctransform, features = c(params$example), vst_outlist = vst_out)
```

```{r, fig.width = 5, fig.height = 10, message = FALSE, warning=FALSE}
wrap_plots(modl)
```

Feature plot as per SCT

```{r, fig.width = 10, fig.height = 5, message = FALSE, warning=FALSE}
wrap_plots(lapply(seurat, plotVariableGenes, selection.method = "sct"), ncol = 2)
```

## Output

Save experiment object:

```{r}
saveRDS(seurat, file = "output/03-normalisation.rds")
```  

Print session information:
  
```{r}
sessionInfo()
```