---
title: "Reduce Dimensions"
params:
  elbow_range:
    - "1:10"
  n.neighbors:
    - 30
  min.dist:
    - 0.3
  perplexity:
    - "N^1/2"
  plot_width:
    - 8
  plot_height:
    - 4
---
  
# Setup

Add child documents

```{r}
#| child = "analysis/big-data.qmd"
```

Restore project dependencies:

```{r}
#renv::restore()
```

Load Bioconductor packages:
  
```{r}
#| message: false
#| warning: false

library(Seurat)
```

Load CRAN packages:
  
```{r}
#| message: false
#| warning: false

library(patchwork)
library(ggplot2)
library(dplyr)
library(reshape2)
library(viridis)
library(plyr)
```

Source user-defined functions:
  
```{r}
#| message: false
#| warning: false

source("scripts/reduceDims.R")
```

# Processing

Read experiment object:
  
```{r}
#| message: false
#| warning: false

seurat <- readRDS("output/03-normalisation.rds")
```

## PCA

PCA individual samples

```{r}
#| message: false
#| warning: false

seurat <- bplapply(seurat, pca, BPPARAM = BPPARAM)
```

### Plot PCAs

Individual PCA - dimplot

```{r}
#| message: false
#| warning: false

pca_plots_dim <- lapply(seurat, dimplot, reduction = "pca", label = FALSE, group.by = "Group")
```

Wrap plot PCAs dimplots

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: PCA plots by group.

wrap_plots(pca_plots_dim, ncol = 2)
```

Individual PCA - featureplot ncount

```{r}
#| message: false
#| warning: false

pca_plots_feature <- lapply(seurat, featureplot, reduction = "pca", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot PCAs featureplot ncount

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: PCA plots by nCount.

wrap_plots(pca_plots_feature, ncol = 2)
```

Individual PCA - featureplot nfeature

```{r}
#| message: false
#| warning: false

pca_plots_feature <- lapply(seurat, featureplot, reduction = "pca", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot PCAs featureplot nfeature

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: PCA plots by nFeature.

wrap_plots(pca_plots_feature, ncol = 2)
```

## Finding dimensionality of dataset

Find ncomponents for downstream dimensionality reduction

### Elbow plot

```{r}
#| message: false
#| warning: false

elbows <- lapply(seurat, elbow)
```

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: Elbow plot.

wrap_plots(elbows, ncol = 2)
```

Take number of PCs from visual inspection - better to go higher

Define elbow vector:

```{r}
#| warning: false

elbow_vector <- params$elbow_range

names(elbow_vector) <- names(seurat)

print(elbow_vector)
```

Add PC elbow to object

```{r}
#| message: false
#| warning: false

seurat <- lapply(seurat, addElbow, elbow_vector = elbow_vector)
```

## UMAP

### Test stability of UMAP params
n.neighbors should be in the range 5 to 50.
min.dist sensible values are in the range 0.001 to 0.5.

```{r}
#| message: false
#| warning: false
#| label: Explore-Features-UMAP

seurat_1 <- bplapply(seurat, umap, n.neighbors = 30, min.dist = 0.3, BPPARAM = BPPARAM)

seurat_2 <- bplapply(seurat, umap, n.neighbors = 5, min.dist = 0.25, BPPARAM = BPPARAM)

seurat_3 <- bplapply(seurat, umap, n.neighbors = 10, min.dist = 0.1, BPPARAM = BPPARAM)

seurat_4 <- bplapply(seurat, umap, n.neighbors = 20, min.dist = 0.1, BPPARAM = BPPARAM)

seurat_5 <- bplapply(seurat, umap, n.neighbors = 10, min.dist = 0.001, BPPARAM = BPPARAM)
```

UMAP n.neighbors = 30, min.dist = 0.3
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP n.neighbors = 30, min.dist = 0.3.

wrap_plots(lapply(seurat_1, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 5, min.dist = 0.25
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP n.neighbors = 5, min.dist = 0.25.

wrap_plots(lapply(seurat_2, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 10, min.dist = 0.1
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP n.neighbors = 10, min.dist = 0.1.

wrap_plots(lapply(seurat_3, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 20, min.dist = 0.1
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP n.neighbors = 20, min.dist = 0.1.

wrap_plots(lapply(seurat_4, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

UMAP n.neighbors = 10, min.dist = 0.001
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP n.neighbors = 10, min.dist = 0.001.

wrap_plots(lapply(seurat_5, dimplot, reduction = "umap", label = FALSE, group.by = "Group"), ncol = 2)
```

Clean up workspace
```{r}
rm(seurat_1, seurat_2, seurat_3, seurat_4, seurat_5)
gc()
```

Run UMAP individual samples for choice of params

```{r}
#| warning: false

print("n.neighbors: ", params$n.neighbors)
print("min.dist: ", params$min.dist)

seurat <- bplapply(seurat, umap, n.neighbors = params$n.neighbors, 
                   min.dist = params$min.dist, BPPARAM = BPPARAM)
```

### Plot UMAP

Individual UMAPs

```{r}
#| message: false
#| warning: false

umap_plots <- lapply(seurat, dimplot, reduction = "umap", label = FALSE, group.by = "Group")
```

Wrap plot UMAP

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP chosen params

wrap_plots(umap_plots, ncol = 2)
```

Individual UMAP - featureplot ncounts

```{r}
#| message: false
#| warning: false

umap_plots_feature <- lapply(seurat, featureplot, reduction = "umap", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot UMAP featureplot ncounts

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP chosen params nCounts

wrap_plots(umap_plots_feature, ncol = 2)
```

Individual UMAP - featureplot nfeatures

```{r}
#| message: false
#| warning: false

umap_plots_feature <- lapply(seurat, featureplot, reduction = "umap", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot UMAP featureplot nfeatures

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: UMAP chosen params nCounts

wrap_plots(umap_plots_feature, ncol = 2)
```

## TSNE

### Test stability of TSNE params

perplexity parameter (should not be bigger than 3 * perplexity < nrow(X) - 1)
NB: Try N^1/2 as rule of thumb

```{r}
#| label: Explore-Features-TSNE-N1-2
#| warning: false
#| message: false

seurat_1 <- bplapply(seurat, tsne, perplexity = "N^1/2", BPPARAM = BPPARAM)
```

```{r}
#| label: Explore-Features-TSNE
#| warning: false
#| message: false

seurat_2 <- bplapply(seurat, tsne, perplexity = 5, BPPARAM = BPPARAM)

seurat_3 <- bplapply(seurat, tsne, perplexity = 10, BPPARAM = BPPARAM)

seurat_4 <- bplapply(seurat, tsne, perplexity = 20, BPPARAM = BPPARAM)

seurat_5 <- bplapply(seurat, tsne, perplexity = 100, BPPARAM = BPPARAM)
```

perplexity = N^1/2
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE perplexity = N^1/2

wrap_plots(lapply(seurat_1, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 5
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE perplexity = 5

wrap_plots(lapply(seurat_2, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 10
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE perplexity = 10

wrap_plots(lapply(seurat_3, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 20
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE perplexity = 20

wrap_plots(lapply(seurat_4, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

perplexity = 100
```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE perplexity = 100

wrap_plots(lapply(seurat_5, dimplot, reduction = "tsne", label = FALSE, group.by = "Group"), ncol = 2)
```

Clean up workspace
```{r}
rm(seurat_1, seurat_2, seurat_3, seurat_4, seurat_5)
gc()
```

Run tsne individual samples for choice of params

```{r}
#| warning: false

print("perplexity: ", params$perplexity)

seurat <- bplapply(seurat, tsne, perplexity = params$perplexity, BPPARAM = BPPARAM)
```

### Plot TSNE

Individual TSNEs

```{r}
#| warning: false
#| message: false

tsne_plots <- lapply(seurat, dimplot, reduction = "tsne", label = FALSE, group.by = "Group")
```

Wrap plot TSNE

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE chosen params.

wrap_plots(tsne_plots, ncol = 2)
```

Individual TSNE - featureplot ncounts

```{r}
#| warning: false
#| message: false

tsne_plots_feature <- lapply(seurat, featureplot, reduction = "tsne", label = FALSE, feature = "nCount_Spatial")
```

Wrap plot TSNE featureplot ncounts

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE chosen params nCounts.

wrap_plots(tsne_plots_feature, ncol = 2)
```

Individual TSNE - featureplot nfeatures

```{r}
#| warning: false
#| message: false

tsne_plots_feature <- lapply(seurat, featureplot, reduction = "tsne", label = FALSE, feature = "nFeature_Spatial")
```

Wrap plot TSNE featureplot nfeatures

```{r}
#| fig.width: params$plot_width
#| fig.height: params$plot_height
#| fig-cap: TSNE chosen params nFeatures.

wrap_plots(tsne_plots_feature, ncol = 2)
```

# Output

Save experiment object:

```{r}
saveRDS(seurat, file = "output/04-reduced-dimensions.rds")
```  

Print session information:
  
```{r}
sessionInfo()
```


