---
title: "Clustering"
author: "Ben Southgate"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: textmate
    df_print: paged
editor_options:
  chunk_output_type: console
---
  
# Setup
  
Set chunk options:
  
```{r}
knitr::opts_chunk$set(echo = TRUE, dev = "png", fig.align = "center")
```

Set library path:
  
```{r}
.libPaths(new = "/sbgenomics/project-files/library")
```

Load Bioconductor packages:
  
```{r, warning=FALSE, message=FALSE}
library(scater)
library(scran)
library(scuttle)
library(Seurat)
```

Load CRAN packages:
  
```{r, warning=FALSE, message=FALSE}
library(patchwork)
library(ggplot2)
library(dplyr)
library(reshape2)
library(clustree)
library(pheatmap)
library(RColorBrewer)
```

Source user-defined functions:
  
```{r, warning=FALSE, message=FALSE}
source("scripts/reduceDims.R")
source("scripts/cluster.R")
```

Enable multicore parallel evaluation:
  
```{r, warning=FALSE, message=FALSE}
library(BiocParallel)

BPP <- MulticoreParam(
  workers = 6,
  stop.on.error = TRUE,
  RNGseed = 0305
)
```

# Processing

Read experiment object:
  
```{r}
seurat <- readRDS("/sbgenomics/output-files/04-reduced-dimensions.rds")
```

## Clustering with SNN

SNN graph construction across a number of params

```{r, warning=FALSE, message=FALSE}
seurat_1 <- bplapply(seurat, snnCluster, k.param = 5, resolution = 0.1, BPPARAM = BPP)
seurat_2 <- bplapply(seurat, snnCluster, k.param = 10, resolution = 0.3, BPPARAM = BPP)
seurat_3 <- bplapply(seurat, snnCluster, k.param = 15, resolution = 0.5, BPPARAM = BPP)
seurat_4 <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.6, BPPARAM = BPP)
seurat_5 <- bplapply(seurat, snnCluster, k.param = 50, resolution = 0.8, BPPARAM = BPP)
```

### Plot umaps

k.param = 5, resolution = 0.1
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_1, dimplot, reduction = "umap", label = T, group.by = "seurat_clusters"), ncol = 2)
```
k.param = 10, resolution = 0.3
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_2, dimplot, reduction = "umap", label = T, group.by = "seurat_clusters"), ncol = 2)
```
k.param = 15, resolution = 0.5
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_3, dimplot, reduction = "umap", label = T, group.by = "seurat_clusters"), ncol = 2)
```
k.param = 20, resolution = 0.6
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_4, dimplot, reduction = "umap", label = T, group.by = "seurat_clusters"), ncol = 2)
```
k.param = 50, resolution = 0.8
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat_5, dimplot, reduction = "umap", label = T, group.by = "seurat_clusters"), ncol = 2)
```
Clean up workspace
```{r}
rm(seurat_1, seurat_2, seurat_3, seurat_4, seurat_5)
gc()
```

### Keeping k stable (saving into seurat object for clustree)

```{r, warning=FALSE, message=FALSE}
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.1, BPPARAM = BPP)
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.2, BPPARAM = BPP)
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.3, BPPARAM = BPP)
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.5, BPPARAM = BPP)
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.6, BPPARAM = BPP)
seurat <- bplapply(seurat, snnCluster, k.param = 20, resolution = 0.8, BPPARAM = BPP)
```
k.param = 20, resolution = 0.1
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.1"), ncol = 2)
```
k.param = 20, resolution = 0.2
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.2"), ncol = 2)
```
k.param = 20, resolution = 0.3
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.3"), ncol = 2)
```

### Plot umaps with k.params 20

k.param = 20, resolution = 0.1
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.1"), ncol = 2)
```

k.param = 20, resolution = 0.2
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.2"), ncol = 2)
```

k.param = 20, resolution = 0.3
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.3"), ncol = 2)
```
k.param = 20, resolution = 0.5
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.5"), ncol = 2)
```

k.param = 20, resolution = 0.6
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.6"), ncol = 2)
```
k.param = 20, resolution = 0.8
```{r, fig.width = 10, fig.height = 10}
wrap_plots(lapply(seurat, dimplot, reduction = "umap", label = T, group.by = "res.0.8"), ncol = 2)
```

## Clustering stability 

Across k = 20 (seurat default)

Using clustree for checking cluster stability across params

```{r, warning=FALSE, message=FALSE}
clustree_plots <- lapply(seurat, clustreeRun)
```

Wrap plot clustree

```{r, fig.width = 12, fig.height = 7}
wrap_plots(clustree_plots[1:3], ncol = 3)
```

```{r, fig.width = 12, fig.height = 7}
wrap_plots(clustree_plots[4:6], ncol = 3)
```

## Silhouette 

Across k = 20 (seurat default)

A cell with a large positive width is closer to other cells from the same cluster compared to cells from different clusters. On the other hand, low or negative widths indicate that cells from different clusters are not well separated.

```{r, warning=FALSE, message=FALSE}
silhouettes_1 <- lapply(seurat, silhouette, clustername = "res.0.2")
silhouettes_2 <- lapply(seurat, silhouette, clustername = "res.0.3")
silhouettes_3 <- lapply(seurat, silhouette, clustername = "res.0.5")
silhouettes_4 <- lapply(seurat, silhouette, clustername = "res.0.6")
silhouettes_5 <- lapply(seurat, silhouette, clustername = "res.0.8")
```

```{r, warning=FALSE, message=FALSE}
silhouette_plots_1 <- lapply(silhouettes_1, plotSil, type = "boxplot", title = "res.0.2")
silhouette_plots_2 <- lapply(silhouettes_2, plotSil, type = "boxplot", title = "res.0.3")
silhouette_plots_3 <- lapply(silhouettes_3, plotSil, type = "boxplot", title = "res.0.5")
silhouette_plots_4 <- lapply(silhouettes_4, plotSil, type = "boxplot", title = "res.0.6")
silhouette_plots_5 <- lapply(silhouettes_5, plotSil, type = "boxplot", title = "res.0.8")
```
Clean up workspace
```{r}
rm(silhouettes_1, silhouettes_2, silhouettes_3, silhouettes_4, silhouettes_5)
gc()
```

#### Wrap plot silhouette

res.0.2
```{r, fig.width = 12, fig.height = 10}
wrap_plots(silhouette_plots_1, ncol = 2)
```
res.0.3
```{r, fig.width = 12, fig.height = 10}
wrap_plots(silhouette_plots_2, ncol = 2)
```
res.0.5
```{r, fig.width = 12, fig.height = 10}
wrap_plots(silhouette_plots_3, ncol = 2)
```
res.0.6
```{r, fig.width = 12, fig.height = 10}
wrap_plots(silhouette_plots_4, ncol = 2)
```
res.0.8
```{r, fig.width = 12, fig.height = 10}
wrap_plots(silhouette_plots_5, ncol = 2)
```

## Spatial feautures 

#### Res 0.2 and 0.3

```{r, warning=FALSE, message=FALSE}
spatial_dim1 <- lapply(seurat, SpatialDimPlot, label = TRUE, label.size = 3, 
group.by = "res.0.2")

spatial_dim2 <- lapply(seurat, SpatialDimPlot, label = TRUE, label.size = 3, 
group.by = "res.0.3")
```

res.0.2
```{r, fig.width = 10, fig.height = 15}
wrap_plots(spatial_dim1, ncol = 2)
```
res.0.3
```{r, fig.width = 10, fig.height = 15}
wrap_plots(spatial_dim2, ncol = 2)
```

## Apply chosen cluster labels

```{r}
seurat <- lapply(seurat, addLabel, chosenCluster = "res.0.3")
```

### UMAP

Individual UMAP

```{r, warning=FALSE}
umap_plots <- lapply(seurat, dimplot, reduction = "umap", group.by = "Cluster")
```

Wrap plot UMAP

```{r, fig.width = 10, fig.height = 10}
wrap_plots(umap_plots, ncol = 2)
```

### TSNE

Individual TSNE

```{r, warning=FALSE}
tsne_plots <- lapply(seurat, dimplot, reduction = "tsne", group.by = "Cluster")
```

Wrap plot TSNE

```{r, fig.width = 10, fig.height = 10}
wrap_plots(tsne_plots, ncol = 2)
```


# Output

Save experiment object:
  
```{r}
saveRDS(seurat, file = "/sbgenomics/output-files/05-clustering.rds")
```  

Print session information:
  
```{r}
sessionInfo()
```

